---
title: Testing the features
output: 
  html_document:
    self_contained: false
    variant: markdown_github
    toc: true
    toc_depth: 2
    preserve_yaml: true
layout: default
---

Plain text

End a line with two spaces
to start a new paragraph.  

*italics* and _italics_

**bold** and __bold__

superscript^2^

~~strikethrough~~

[link](http://peter.solymos.org)

# Header 1

## Header 2

### Header 3

#### Header 4

##### Header 5

###### Header 6

endash: --

emdash: ---

ellipsis: ...

inline equation: $A = \pi*r^{2}$

$$A = \pi*r^{2}$$

image: ![](favicon.ico)

horizontal rule (or slide break):

***

> block quote

* unordered list
* item 2
 + sub-item 1
 + sub-item 2

1. ordered list
2. item 2
 + sub-item 1
 + sub-item 2

Table Header | Second Header
------------- | -------------
Table Cell | Cell 2
Cell 3 | Cell 4 

Make a code chunk with three back ticks followed
by an r in braces. End the chunk with three back
ticks: 

```{r}
paste("Hello", "World!")
``` 

Place code inline with a single back ticks. The
first back tick must be followed by an R, like
this `r paste("Hello", "World!")`.

Add chunk options within braces. For example,
`echo=FALSE` will prevent source code from being
displayed:

```{r eval=TRUE, echo=FALSE}
paste("Hello", "World!")
``` 


# Data cloning

## Here is a website

Paragraph text.

Some MathJax: inline eq $$ x_{i}=exp(\lambda^2_{i}) $$ like this.

Or:

$$ a = \sqrt{b^2 + c^2} $$


Some MathJax: \\(x_{i}=exp(\lambda^2_{i}))\\

Or:

\\[ a = \sqrt{b^2 + c^2} ]\\

This is inline `code` and this is code block:

```R
## with some notes
a <- 1 + 2 # here
(b <- a * 2)
```

Well.

Lists:

1. One,
2. Two.

And:

* item
* item
  and beyond

finally GFM checkbox:

- [ ] todo 1
- [x] done,
- [ ] not.

> We're living the future so
> the present is our past.


First Header | Second Header
------------ | -------------
Content from cell 1 | Content from cell 2
Content in the first column | Content in the second column

An ~~this is strikeout~~ text.

Is it?
