---
title: "Abundance models with detection error"
author: "Peter Solymos and Subhash Lele"
date: "July 16, 2016 &mdash; Madison, WI &mdash; NACCB Congress"
layout: course
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
course:
  location: Madison
  year: 2016
  title: "Hierarchical Models Made Easy &mdash; July 16, 2016 &mdash; Madison, WI &mdash; NACCB Congress"
  lecture: Abundance models with detection error
  file: n-mixture
  previous: occupancy
  next: lmm
  pdf: n-mixture.pdf
---

We can easily generalize this to model abundance surveys. The N-mixture model is the simplest (though unrealistic in practice). 

### Assumptions

* Replicate surveys, 
* independence, 
* closed population.

### Specification of the hierarchical model

* True abundance model:  $$ N_i \sim Poisson(\lambda) $$ for locations $i=1,2, \ldots, n$. 
* Observation model: $$ (Y_{i,t} \mid N_i) \sim Binomial(N_i, p) $$ for visits $t=1,2, \ldots, T$.

```{r}
set.seed(1234)
n <- 200
T <- 1
p <- 0.6
lambda <- 4.2
N <- rpois(n = n, lambda = lambda)
Y <- matrix(NA, n, T)
for (t in 1:T) {
    Y[,t] <- rbinom(n = n, size = N, prob = p)
}
table(N = N, Y = apply(Y, 1, max))
```

```{r}
library(dclone)
library(rjags)
model <- custommodel("model {
    for (i in 1:n) {
        N[i] ~ dpois(lambda)
        for (t in 1:T) {
            Y[i,t] ~ dbin(p, N[i])
        }
    }
    p ~ dunif(0.001, 0.999)
    lambda ~ dlnorm(0, 0.001)
}")
dat <- list(Y = Y, n = n, T = T)
ini <- list(N = apply(Y, 1, max) + 1)
fit <- jags.fit(data = dat, params = c("p", "lambda"), 
    n.update = 10000,
    model = model, inits = ini)
summary(fit)
gelman.diag(fit)
plot(fit)
pairs(fit)
```

```{r}
ifun <- function(model, n.clones) {
    dclone(list(N = apply(Y, 1, max) + 1), n.clones)
}
dcfit <- dc.fit(data = dat, 
    params = c("p", "lambda"), model = model,
    inits = ini, initsfun = ifun,
    n.clones = c(1, 2, 4, 8), 
    unchanged = "T", multiply = "n")
summary(dcfit)
plot(dcfit)
dctable(dcfit)
plot(dctable(dcfit))
dcdiag(dcfit)
plot(dcdiag(dcfit))
pairs(dcfit)
```

As before it is easy to include covariates in the models. There are various extensions and modifications proposed to this basic model. See Lele et al., Solymos et al, Solymos and Lele, Dail and Madsen. (Here we can advertise our work on single survey method and the poster.)

Single visit abundance model with covariates:

```{r}
set.seed(1234)
n <- 200
x <- rnorm(n)
z <- rnorm(n)
beta <- c(0.9, 0.5)
theta <- c(0.8, -0.5)
Z <- model.matrix(~z)
X <- model.matrix(~x)
p <- plogis(Z %*% theta)
lambda <- exp(X %*% beta)
N <- rpois(n = n, lambda = lambda)
Y <- rbinom(n = n, size = N, prob = p)
table(N = N, Y = Y)
## naive abundance parameter estimates
m <- glm(Y ~ x, family = poisson("log"))
coef(m)

library(detect)
md <- svabu(Y ~ x | z, zeroinfl = FALSE)
coef(md)
```

```{r}
model <- custommodel("model {
    for (i in 1:n) {
        N[i] ~ dpois(lambda[i])
        Y[i] ~ dbin(p[i], N[i])
        log(lambda[i]) <- inprod(X[i,], beta)
        logit(p[i]) <- inprod(Z[i,], theta)
    }
    for (j in 1:px) {
        beta[j] ~ dnorm(naive[j], 0.1)
    }
    for (j in 1:pz) {
        theta[j] ~ dnorm(0, 0.01)
    }
}")
dat <- list(Y = Y, n = n, X = X, Z = Z, 
    px = ncol(X), pz = ncol(Z), naive = coef(m))
ini <- list(N = Y + 1)
fit <- jags.fit(data = dat, params = c("beta", "theta"), 
    n.update = 5000,
    model = model, inits = ini)
## DC
ifun <- function(model, n.clones) {
    dclone(list(N = Y + 1), n.clones)
}
dcfit <- dc.fit(data = dat, 
    params = c("beta", "theta"), model = model,
    inits = ini, initsfun = ifun,
    n.clones = c(1, 2, 4, 8), 
#    n.update = 5000,
    unchanged = c("px", "pz", "naive"), multiply = "n")
```

## Learning with DC

```{r}
model <- custommodel("model {
    for (i in 1:n) {
        N[i] ~ dpois(lambda[i])
        Y[i] ~ dbin(p[i], N[i])
        log(lambda[i]) <- inprod(X[i,], beta)
        logit(p[i]) <- inprod(Z[i,], theta)
    }

    cf[1:(px + pz)] ~ dmnorm(pr[,1], pr[,2:(px + pz + 1)])
    beta <- cf[1:px]
    theta <- cf[(px + 1):(px + pz)]
}")
dat <- list(Y = Y, n = n, X = X, Z = Z, 
    px = ncol(X), pz = ncol(Z), 
    pr = unname(cbind(c(coef(m), rep(0, ncol(Z))),
        diag(0.01, ncol(X) + ncol(Z)))))
ini <- list(N = Y + 1)
ifun <- function(model, n.clones) {
    dclone(list(N = Y + 1), n.clones)
}
ufun <- function(model, n.clones) {
    cbind(coef(model), solve(vcov(model)))
}
dcfit <- dc.fit(data = dat, 
    params = c("beta", "theta"), model = model,
    inits = ini, initsfun = ifun, 
    update = "pr", updatefun = ufun,
    n.clones = c(1, 2, 4, 8), 
    n.update = 5000,
    unchanged = c("px", "pz", "pr"), multiply = "n")
```

### Zero-inflated Poisson latent process

This really becomes an issue when $T = 1$.
With $T > 1$ it is much easier to distinguish
non occupied ($O_i = 0$ or $N_i = 0 | O_i = 1$) locations
when all the detection history is 0,
and non-detections when some of the detection history is >0
if $p$ is not too small.

```{r}
set.seed(1234)
n <- 100
T <- 2
p <- 0.6
lambda <- 3.5
q <- 0.25
O <- rbinom(n, size = 1, prob = q)
N <- O * rpois(n = n, lambda = lambda)
Y <- matrix(NA, n, T)
for (t in 1:T) {
    Y[,t] <- rbinom(n = n, size = N, prob = p)
}
table(N = N, Y = apply(Y, 1, max))
```

```{r}
library(dclone)
model <- custommodel("model {
    for (i in 1:n) {
        O[i] ~ dbern(q)
        N[i] ~ dpois(lambda * O[i])
        for (t in 1:T) {
            Y[i,t] ~ dbin(p, N[i])
        }
    }
    p ~ dunif(0.001, 0.999)
    lambda ~ dlnorm(0, 0.001)
    q ~ dunif(0.001, 0.999)
}")
dat <- list(Y = Y, n = n, T = T)
ini <- list(N = ifelse(rowSums(Y) > 0, 1, 0) * (apply(Y, 1, max) + 1),
    O = ifelse(rowSums(Y) > 0, 1, 0))
fit <- jags.fit(data = dat, params = c("p", "lambda", "q"), 
    n.update = 10000,
    model = model, inits = ini)
summary(fit)
gelman.diag(fit)
plot(fit)
pairs(fit)
```

